Experimental and numerical studies of the effects of resonant and random excitations of the proton beam in the LHC, with applications to the design of a pulsed hollow electron lens for active halo control

### 2018-03-14 resex.tex
Revised results of 7th-turn pulsing, including figures and captions. Cropped 3 plots. Started using abbreviated commands for n-th turn pulsing.

\item \textbf{resonant excitation:} The electron beam is switched on
  only every $k^{\mathrm{th}}$ turn. The excitation can be represented
  by the following expression:
  \begin{eqnarray}
    \label{intro:eqn:1}
    f(t) & = & \sum_{n=-\infty}^{+\infty}\delta\left[t-n\cdot(kT)\right],
  \end{eqnarray}
  where $n$ is the turn number and $T$ is the revolution period. Its
  Fourier representation is
  \begin{equation}
    \label{intro:eqn:2}
    f(t) = \Sh_{kT}(t) = \frac{1}{kT}\sum_{n=-\infty}^{+\infty}e^{2\pi i f_nt},
  \end{equation}
  with $f_n = n \cdot f_{\rm rev} / k$ and
  where $\Sh_{kT}$ is the Dirac comb. In general,
  $k^{\mathrm{th}}$~turn pulsing drives $k^{\mathrm{th}}$-order
  resonances~\cite{md_sim_hel_res_ex_fitterer}. This type of pulsing
  pattern was used in the Tevatron during regular collider operations
  for abort-gap cleaning~\cite{hel_tevatron_abortgap_zhang}.
\end{itemize}

@@

\begin{table*}
  \caption{Beam parameters and machine configuration for the two
    resonant excitation experiments of~2016
    and~2017~\cite{resexmd2016, resexmd2017}. The plane of the
    excitation is abbreviated as H for horizontal, V for vertical and
    H+V for horizontal and vertical at the same time.}
  \label{tab:md_param}
  \begin{ruledtabular}
    \begin{tabular}{lcc}
      Parameter & Experiment 2016 & Experiment 2017  \\
      \colrule
      beam &\multicolumn{2}{c}{Beam~1} \\
      beam energy &\multicolumn{2}{c}{injection energy, 450 GeV} \\\hline
      single bunch intensity &\multicolumn{2}{c}{$0.7\times10^{11}$} \\
      normalized emittance &\multicolumn{2}{c}{$2.5$--\q{3.5}{\mu m}} \\
      $4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns}\\
      $1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm}\\
      number of bunches & $12 \times 4 = 48$ & $3 \times 72 = 216$ \\
                &  & (+ 1 pilot + 12 nominal) \\\hline
      injection optics, $\beta^* = \q{11}{m}$ & standard optics 2016 & standard optics 2017\\
      Landau-damping octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MO}}
                                  = \q{+19.6}{A}$ for MOF circuit and}\\
                & \multicolumn{2}{c}{$I_{\mathrm{MO}} = \q{-19.6}{A}$
                  for MOD circuit (standard 2016 settings)} \\ \hline
      working point $(Q_x, Q_y)$ & (64.28, 59.31) & (62.27, 60.295) \\
      chromaticity $(Q'_x, Q'_y)$ & \multicolumn{2}{c}{(+15, +15)}\\ \hline
      pulsing patterns  & $7^{\mathrm{th}}$~turn H
                                  &$7^{\mathrm{th}}$~turn H, V, H+V \\
                & $10^{\mathrm{th}}$~turn V & $8^{\mathrm{th}}$~turn
                                              H, V, H+V \\
                & &  random  H, V, H+V\\
    \end{tabular}
  \end{ruledtabular}
\end{table*}

@@

The choice of excitation patterns for the experiments was guided by
losses and emittance growths calculated in numerical tracking
simulations, described below. It was chosen to study experimentally
the two pulsing patterns with the largest calculated effects on the
beam ($7^{\mathrm{th}}$- and $10^{\mathrm{th}}$-turn pulsing), one
pattern with no effect ($8^{\mathrm{th}}$-turn pulsing), and the
random excitation. In order to quantify the reproducibility of the
results under different machine configurations, one pulsing pattern
($7^{\mathrm{th}}$-turn pulsing) was tried first in~2016 and then
repeated in~2017.

@@

In this Section, we present the experimental observations and compare
them with the results of numerical simulations.  In
Section~\ref{sec:pattern}, we show which pulsing patterns are
predicted to be the most efficient in the LHC.  Next, we present
results for each of the specific excitation patterns that could be
tested experimentally, namely pulsing every $10^{\mathrm{th}}$~turn
(Section~\ref{sec:simex10}), $7^{\mathrm{th}}$~turn
(Section~\ref{sec:simex7}), $8^{\mathrm{th}}$~turn
(Section~\ref{sec:simex8}), and the random excitation
(Section~\ref{sec:simexran}).  In Section~\ref{sec:damp}, we describe
how the transverse damping system influenced the effects of the
external excitation sources on losses, beam distributions, and
emittances.

@@

\begin{figure*}
  \begin{tabular}{cc}
    no excitation & $10^{\mathrm{th}}$, H+V \\
    \includegraphics[width=0.49\textwidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png} &
    \includegraphics[width=0.49\textwidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
    \\
    $7^{\mathrm{th}}$, H & $7^{\mathrm{th}}$, V \\
    \includegraphics[width=0.49\textwidth]{2016injnocolc15o+19_6noerrut7skh_dp0_ord7.png} &
    \includegraphics[width=0.49\textwidth]{2016injnocolc15o+19_6noerrut7skv_dp0_ord7.png} \\
  \end{tabular}
  \caption{Frequency-map analysis in betatron tune space based on the
    2016 injection optics with no machine errors and tunes (64.28,
    59.31): without excitation (top left); $10^{\mathrm{th}}$~turn
    pulsing in both horizontal and vertical planes (top left);
    $7^{\mathrm{th}}$~turn pulsing in the horizontal (bottom left) and
    vertical (bottom right) planes. The excitation amplitude is
    120~nrad in the corresponding plane. The colors (blue to red)
    represent the tune jitter of tracked particles starting at each
    given location in tune space~\cite{fmalaskar}. The absence of a
    strong excitation of any resonance in case of vertical
    $7^{\mathrm{th}}$~turn pulsing and the strong excitation in case
    of horizontal pulsing confirms the excitation of the $7 Q_x$
    resonance. For $10^{\mathrm{th}}$~turn pulsing, there is in
    contrast no significant difference between H, V, or H+V (see
    Appendix~\ref{app:sec:10}, Fig.~\ref{app:fig:fma:10}).}
  \label{fig:patternfma}
\end{figure*}

\begin{figure*}
  \begin{tabular}{cc}
    no excitation & random, H+V \\
    \includegraphics[width=0.49\textwidth]{2016injerra2b2u_3_5um_hist_y.png} &
    \includegraphics[width=0.49\textwidth]{2016injerra2b2u_ranhv_3_5um_hist_x.png} \\
    $7^{\mathrm{th}}$, H+V & $10^{\mathrm{th}}$, H+V \\
    \includegraphics[width=0.49\textwidth]{2016injerra2b2u_t7skhv_3_5um_hist_x.png}
                  &
    \includegraphics[width=0.49\textwidth]{2016injerra2b2u_t10skhv_3_5um_hist_y.png} \\
  \end{tabular}
  \caption{Calculated beam distributions as a function of vertical
    position from distribution-tracking simulations based on the 2016
    injection optics with $(Q_x, Q_y) = (64.28, 59.31)$ and standard
    lattice errors: no excitation (top left), random excitation (top
    right), $7^{\mathrm{th}}$~turn pulsing (bottom left), and
    $10^{\mathrm{th}}$~turn pulsing (bottom right). The excitations
    are applied in both planes with an amplitude of 96~nrad. For each
    of the 4~cases, 3~plots are shown. The top plot shows the
    normalized transverse distributions: `initial' (after
    $10^2$~turns, in gray), `final' (after $10^4$~turns, in black),
    and their Gaussian fits (light and dark red, respectively). The
    middle plots show the relative residuals (i.e., differences, in
    percent of the peak value) between final and initial distributions
    (in black) and between each distribution and its Gaussian fit (in
    light and dark red). The ratios between final and initial
    distributions are drawn in black in the bottom plots. The
    residuals emphasize changes near the core of the distributions,
    whereas ratios (when statistically significant), reveal variations
    in the tails.}
  \label{fig:patternhist}
\end{figure*}

@@

As an example, the simulation results for the 2016 experiment are
shown in Fig.~\ref{fig:patternsim}.  A clear dependence of both losses
and emittances on the pattern can be seen.  The largest losses are
predicted for $3^{\mathrm{rd}}$, $7^{\mathrm{th}}$ and
$10^{\mathrm{th}}$~turn pulsing, and for a uniform random
excitation. Because the bunch length decreases with the number of lost
particles, losses are attributed to off-momentum particles hitting the
transverse aperture. Significant emittance growth is visible only for
$7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing and for the
random excitation. Compared to any resonant excitation, the random
excitation shows by far the strongest effect.

The effects of $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing
are also observed in the absence of machine lattice errors. In this
case, the only sources of nonlinearity are sextupoles and
octupoles~\cite{md_sim_hel_res_ex_fitterer}, suggesting that these
nonlinearities are responsible for the pronounced beam
sensitivity. The driven resonances are revealed by the frequency-map
analysis, shown in Fig.~\ref{fig:patternfma}. The $7 Q_x$ resonance is
excited by the $7^{\mathrm{th}}$~turn pulsing, and the $10 Q_x$ and
$10 Q_y$ resonances by the $10^{\mathrm{th}}$~turn pulsing. As
octupoles can only drive even resonances, the sources of the $7 Q_x$
resonances are the sextupoles, while the octupoles generate the tune
footprint. The other pulsing patterns do not exhibit any increase in
losses or emittance growth without magnetic errors. Their effect can
be thus attributed to an interaction of the excitation with the
lattice errors, implying also a sensitivity to the chosen random seed
in simulations and to the specific lattice in experiments.

In Fig.~\ref{fig:patternsim}, one can see that, in the cases of
$7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing, the emittances
start from an increased initial value and then stay almost constant
for the duration of the simulation. This behavior is typical of the
resonant excitation and it is not an artifact of the simulation. It
was verified that it is due to the adjustment of the input beam
distribution to a new equilibrium during the first $10^4$~turns. In
all long-term simulations ($10^6$~turns, corresponding to 89~s)
presented in this paper, the beam distribution is saved every
$10^4$~turns. Therefore, this initial fast adjustment manifests itself
in an increased initial emittance value.

Besides having a larger emittance, the new distribution also differs
from a Gaussian distribution.  This is illustrated in
Fig.~\ref{fig:patternhist}, which shows, for different excitation
patterns, two snapshots of the normalized transverse beam
distributions, the `initial' one at $10^2$~turns and the `final' one
at $10^4$~turns, and how they differ from each other and from a
Gaussian distribution. For instance, for $7^{\mathrm{th}}$ turn
pulsing (bottom left plot in Fig.~\ref{fig:patternhist}), simulations
indicate a clear shift in beam population from the core ($< 1\sigma$)
towards the regions at $\pm 1.5\sigma$ and towards the tails at
$\pm 4\sigma$.

@@

\begin{figure*}
  \begin{tabular}{cc}
    Reference bunch, no excitation & Excited bunch, maximum amplitude \\
    \includegraphics[width=0.48\textwidth]{profile_v_10thv_slot_50.png} &
    \includegraphics[width=0.48\textwidth]{profile_v_10thv_slot_1300.png}\\
    \end{tabular}
    \caption{Vertical beam profiles measured with the Beam Synchrotron
      Radiation Telescope (BSRT) during the 2016 experiments on
      $10^{\mathrm{th}}$~turn pulsing in the vertical plane. The
      transverse damping system was not active on these bunches. The
      beam distributions at the end of the excitation period are shown
      in black in the top plots, with a Gaussian fit in red. The
      bottom plots show the residuals: final profile minus initial
      profile (black); final profile minus its Gaussian fit
      (red). Residuals are expressed as a fraction of the peak profile
      amplitude. The black lines in the plots of the residuals are a
      measure of the overall change of the distribution shape. The red
      lines indicate how different the final distributions are from a
      Gaussian shape. Details of the analysis are given in
      Ref.~\cite{bsrtprofinj}. The distribution of the reference bunch
      (left) is almost unchanged, whereas the bunch experiencing the
      maximum excitation (right) shows a clear shift of particles from
      the axis towards approximately $\pm 2\sigma$.}
    \label{fig:10thexpprof}
\end{figure*}

\begin{figure*}
  \begin{tabular}{cc}
    \includegraphics[width=0.49\textwidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_intensity.png} &
    \includegraphics[width=0.49\textwidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2_rel.png} \\
    \includegraphics[width=0.49\textwidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit1.png} &
    \includegraphics[width=0.49\textwidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2.png} \\
  \end{tabular}	
  \caption{Simulations (with the distribution-tracking method) based
    on the 2016 injection optics with standard lattice errors and
    tunes $(Q_x, Q_y) = (64.28, 59.31)$: relative beam intensity (top
    left); normalized horizontal (bottom left) and vertical (bottom
    right) emittances; vertical emittance relative to its initial
    value (top right). The solid lines indicate the effect of
    $10^{\mathrm{th}}$~turn pulsing in the vertical plane. The dotted
    lines include both the resonant excitation and a random dipole
    noise component of 6~nrad (in both horizontal and vertical
    planes).}
  \label{fig:10thsim}
\end{figure*}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{2016_scale_amp_10v_ran_lblshort_sim_prstab.png}
  \caption{Comparison between experiments and simulations of loss
    rates vs.\ resonant excitation amplitude for
    $10^{\mathrm{th}}$~turn pulsing in the vertical plane. The
    relative average loss rate~$R$ is defined in
    Eq.~\ref{eqn:lossrate}. The experimental results for a maximum
    excitation amplitude of 48~nrad are shown in black and those for
    96~nrad are plotted in green (see also Fig.~\ref{fig:10thexp},
    left, for instance). The results with damper off are represented
    by solid dots, those with damper on are shown with open
    circles. The simulation results, including random dipole noise,
    are shown in red (see also Fig.~\ref{fig:10thsim}). Statistical
    and systematic uncertainties are discussed in the text. The curves
    represent empirical quadratic fits to the data with damper off
    (solid) and with damper on (dashed).}
  \label{fig:10thexploss}
\end{figure}

The $10^{\mathrm{th}}$~turn pulsing pattern was tested in 2016 with an
excitation in the vertical plane only. As this was the first time a
resonant excitation was tested in the LHC, only 48~bunches were
injected, in order to guarantee safe operation of the machine. To test
different excitation amplitudes during one fill and have enough
statistics, the excitation was only applied in one plane. The
experiment started with a period of 12~minutes without excitation, to
let the beam distribution fully adjust to its equilibrium state after
injection. The excitation was then applied for 11~minutes, with the
excitation scheme shown in Fig.~\ref{fig:fill} and a maximum
excitation amplitude of
$A_{\mathrm{max}} = 5 \, \Delta A = \q{48}{nrad}$. The excitation was
then further increased to the maximum value of
$A_{\mathrm{max}} = 5 \, \Delta A = \q{96}{nrad}$ and kept for another
period of 11~minutes.

@@

Numerical simulations of intensities and emittances for this
experiment are shown in Fig.~\ref{fig:10thsim}. They are the results
of distribution tracking for $10^{\mathrm{th}}$~turn pulsing without
additional noise (solid lines) and with an additional noise component
(dotted lines), to emulate the natural noise present in the LHC
(Eq.~\ref{eq:noise-kick}). The presence of noise significantly changes
the calculated effect of the resonant excitation on both losses and
emittances.

@@

% \begin{figure*}
% 		2016 experiment, $7^\mathrm{th}$ H	\\	\begin{minipage}[t]{0.32\linewidth}
% 		\centering
% 		\includegraphics[height=0.75\linewidth]{2016_bunch_intensity_h7th_no_damper_avg.png}
% 	\end{minipage}	
% 	\begin{minipage}[t]{0.32\linewidth}
% 		\centering
% 		\includegraphics[height=0.75\linewidth]{2016_emith_avg_rel_h7th_no_damper.png}
% 	\end{minipage}	
% 	\begin{minipage}[t]{0.32\linewidth}
% 		\centering
% 		\includegraphics[height=0.75\linewidth]{2016_emitv_avg_rel_h7th_no_damper.png}
% 	\end{minipage}	
% 		2017 experiment, $7^\mathrm{th}$ H+V\\
% 	\begin{minipage}[t]{0.32\linewidth}
% 		\centering
% 		\includegraphics[height=0.75\linewidth]{2017_bunch_intensity_hv7th_no_damper_avg.png}
% 	\end{minipage}	
% 	\begin{minipage}[t]{0.32\linewidth}
% 		\centering
% 		\includegraphics[height=0.75\linewidth]{2017_emith_avg_rel_hv7th_no_damper.png}
% 	\end{minipage}	
% 	\begin{minipage}[t]{0.32\linewidth}
% 		\centering
% 		\includegraphics[height=0.75\linewidth]{2017_emitv_avg_rel_hv7th_no_damper.png}
% 	\end{minipage}	
% 	\caption{\label{fig:7thexp} Relative horizontal (top) and vertical (bottom) emittance obtained during the 2016 experiments (left) and the 2017 experiment (right) averaged over the bunches experiencing the same excitation amplitude. For all bunches the damper was not active.}
% \end{figure*}

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{2016+2017_scale_amp_7h_lblshort.png}
	\caption{\label{fig:7thexploss} Comparison of scaling of loss rates with excitation amplitude for $7^{\mathrm{th}}$~turn pulsing in H and data obtained during the 2016 experiment (black and blue) and 2017 experiment (red and yellow). The excitation amplitude error is 50\% in 2016 and was reduced to 15\% in 2017. The solid and dashed line indicate the second order polynomial fit to the loss rate.}
\end{figure}

>>>>

\item \textbf{resonant excitation:} The electron beam is switched on
  only every $k$-th turn. The excitation can be represented
  by the following expression:
  \begin{eqnarray}
    \label{intro:eqn:1}
    f(t) & = & \sum_{n=-\infty}^{+\infty}\delta\left[t-n\cdot(kT)\right],
  \end{eqnarray}
  where $n$ is the turn number and $T$ is the revolution period. Its
  Fourier representation is
  \begin{equation}
    \label{intro:eqn:2}
    f(t) = \Sh_{kT}(t) = \frac{1}{kT}\sum_{n=-\infty}^{+\infty}e^{2\pi i f_nt},
  \end{equation}
  with $f_n = n \cdot f_{\rm rev} / k$ and where $\Sh_{kT}$ is the
  Dirac comb. In general, $k$-th~turn pulsing drives $k$-th order
  resonances~\cite{md_sim_hel_res_ex_fitterer}. This type of pulsing
  pattern was used in the Tevatron during regular collider operations
  for abort-gap cleaning~\cite{hel_tevatron_abortgap_zhang}.
\end{itemize}

@@

\begin{table*}
  \caption{Beam parameters and machine configuration for the two
    resonant excitation experiments of~2016
    and~2017~\cite{resexmd2016, resexmd2017}. The plane of the
    excitation is abbreviated as H for horizontal, V for vertical and
    H+V for horizontal and vertical at the same time.}
  \label{tab:md_param}
  \begin{ruledtabular}
    \begin{tabular}{lcc}
      Parameter & Experiment 2016 & Experiment 2017  \\
      \colrule
      beam &\multicolumn{2}{c}{Beam~1} \\
      beam energy &\multicolumn{2}{c}{injection energy, 450 GeV} \\\hline
      single bunch intensity &\multicolumn{2}{c}{$0.7\times10^{11}$} \\
      normalized emittance &\multicolumn{2}{c}{$2.5$--\q{3.5}{\mu m}} \\
      $4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns}\\
      $1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm}\\
      number of bunches & $12 \times 4 = 48$ & $3 \times 72 = 216$ \\
                &  & (+ 1 pilot + 12 nominal) \\\hline
      injection optics, $\beta^* = \q{11}{m}$ & standard optics 2016 & standard optics 2017\\
      Landau-damping octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MO}}
                                  = \q{+19.6}{A}$ for MOF circuit and}\\
                & \multicolumn{2}{c}{$I_{\mathrm{MO}} = \q{-19.6}{A}$
                  for MOD circuit (standard 2016 settings)} \\ \hline
      working point $(Q_x, Q_y)$ & (64.28, 59.31) & (62.27, 60.295) \\
      chromaticity $(Q'_x, Q'_y)$ & \multicolumn{2}{c}{(+15, +15)}\\ \hline
      pulsing patterns  & 7th~turn H
                                  &7th~turn H, V, H+V \\
                & $10^{\mathrm{th}}$~turn V & $8^{\mathrm{th}}$~turn
                                              H, V, H+V \\
                & &  random  H, V, H+V\\
    \end{tabular}
  \end{ruledtabular}
\end{table*}

@@

The choice of excitation patterns for the experiments was guided by
losses and emittance growths calculated in numerical tracking
simulations, described below. It was chosen to study experimentally
the two pulsing patterns with the largest calculated effects on the
beam (7th- and $10^{\mathrm{th}}$-turn pulsing), one pattern with no
effect ($8^{\mathrm{th}}$-turn pulsing), and the random excitation. In
order to quantify the reproducibility of the results under different
machine configurations, one pulsing pattern (\seventhtp) was tried
first in~2016 and then repeated in~2017.

@@

In this Section, we present the experimental observations and compare
them with the results of numerical simulations.  In
Section~\ref{sec:pattern}, we show which pulsing patterns are
predicted to be the most efficient in the LHC.  Next, we present
results for each of the specific excitation patterns that could be
tested experimentally, namely pulsing every $10^{\mathrm{th}}$~turn
(Section~\ref{sec:simex10}), 7th~turn
(Section~\ref{sec:simex7}), $8^{\mathrm{th}}$~turn
(Section~\ref{sec:simex8}), and the random excitation
(Section~\ref{sec:simexran}).  In Section~\ref{sec:damp}, we describe
how the transverse damping system influenced the effects of the
external excitation sources on losses, beam distributions, and
emittances.

@@

\begin{figure*}
  \begin{tabular}{cc}
    no excitation & $10^{\mathrm{th}}$, H+V \\
    \includegraphics[width=0.49\textwidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png} &
    \includegraphics[width=0.49\textwidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
    \\
    7th, H & 7th, V \\
    \includegraphics[width=0.49\textwidth]{2016injnocolc15o+19_6noerrut7skh_dp0_ord7.png} &
    \includegraphics[width=0.49\textwidth]{2016injnocolc15o+19_6noerrut7skv_dp0_ord7.png} \\
  \end{tabular}
  \caption{Frequency-map analysis in betatron tune space based on the
    2016 injection optics with no machine errors and tunes (64.28,
    59.31): without excitation (top left); $10^{\mathrm{th}}$~turn
    pulsing in both horizontal and vertical planes (top left);
    \seventhtp\ in the horizontal (bottom left) and
    vertical (bottom right) planes. The excitation amplitude is
    120~nrad in the corresponding plane. The colors (blue to red)
    represent the tune jitter of tracked particles starting at each
    given location in tune space~\cite{fmalaskar}. The absence of a
    strong excitation of any resonance in case of vertical
    \seventhtp\ and the strong excitation in case
    of horizontal pulsing confirms the excitation of the $7 Q_x$
    resonance. For $10^{\mathrm{th}}$~turn pulsing, there is in
    contrast no significant difference between H, V, or H+V (see
    Appendix~\ref{app:sec:10}, Fig.~\ref{app:fig:fma:10}).}
  \label{fig:patternfma}
\end{figure*}

\begin{figure*}
  \begin{tabular}{cc}
    no excitation & random, H+V \\
    \includegraphics[width=0.49\textwidth]{2016injerra2b2u_3_5um_hist_y.png} &
    \includegraphics[width=0.49\textwidth]{2016injerra2b2u_ranhv_3_5um_hist_x.png} \\
    7th, H+V & $10^{\mathrm{th}}$, H+V \\
    \includegraphics[width=0.49\textwidth]{2016injerra2b2u_t7skhv_3_5um_hist_x.png}
                  &
    \includegraphics[width=0.49\textwidth]{2016injerra2b2u_t10skhv_3_5um_hist_y.png} \\
  \end{tabular}
  \caption{Calculated beam distributions as a function of vertical
    position from distribution-tracking simulations based on the 2016
    injection optics with $(Q_x, Q_y) = (64.28, 59.31)$ and standard
    lattice errors: no excitation (top left), random excitation (top
    right), \seventhtp\ (bottom left), and
    \tenthtp\ (bottom right). The excitations
    are applied in both planes with an amplitude of 96~nrad. For each
    of the 4~cases, 3~plots are shown. The top plot shows the
    normalized transverse distributions: `initial' (after
    $10^2$~turns, in gray), `final' (after $10^4$~turns, in black),
    and their Gaussian fits (light and dark red, respectively). The
    middle plots show the relative residuals (i.e., differences, in
    percent of the peak value) between final and initial distributions
    (in black) and between each distribution and its Gaussian fit (in
    light and dark red). The ratios between final and initial
    distributions are drawn in black in the bottom plots. The
    residuals emphasize changes near the core of the distributions,
    whereas ratios (when statistically significant), reveal variations
    in the tails.}
  \label{fig:patternhist}
\end{figure*}

@@

As an example, the simulation results for the 2016 experiment are
shown in Fig.~\ref{fig:patternsim}.  A clear dependence of both losses
and emittances on the pattern can be seen.  The largest losses are
predicted for $3^{\mathrm{rd}}$, 7th and
$10^{\mathrm{th}}$~turn pulsing, and for a uniform random
excitation. Because the bunch length decreases with the number of lost
particles, losses are attributed to off-momentum particles hitting the
transverse aperture. Significant emittance growth is visible only for
7th- and \tenthtp\ and for the
random excitation. Compared to any resonant excitation, the random
excitation shows by far the strongest effect.

The effects of 7th- and \tenthtp\ are also observed in the absence of
machine lattice errors. In this case, the only sources of nonlinearity
are sextupoles and octupoles~\cite{md_sim_hel_res_ex_fitterer},
suggesting that these nonlinearities are responsible for the
pronounced beam sensitivity. The driven resonances are revealed by the
frequency-map analysis, shown in Fig.~\ref{fig:patternfma}. The
$7 Q_x$ resonance is excited by the \seventhtp, and the $10 Q_x$ and
$10 Q_y$ resonances by the $10^{\mathrm{th}}$~turn pulsing. As
octupoles can only drive even resonances, the sources of the $7 Q_x$
resonances are the sextupoles, while the octupoles generate the tune
footprint. The other pulsing patterns do not exhibit any increase in
losses or emittance growth without magnetic errors. Their effect can
be thus attributed to an interaction of the excitation with the
lattice errors, implying also a sensitivity to the chosen random seed
in simulations and to the specific lattice in experiments.

In Fig.~\ref{fig:patternsim}, one can see that, in the cases of 7th-
and \tenthtp, the emittances start from an increased initial value and
then stay almost constant for the duration of the simulation. This
behavior is typical of the resonant excitation and it is not an
artifact of the simulation. It was verified that it is due to the
adjustment of the input beam distribution to a new equilibrium during
the first $10^4$~turns. In all long-term simulations ($10^6$~turns,
corresponding to 89~s) presented in this paper, the beam distribution
is saved every $10^4$~turns. Therefore, this initial fast adjustment
manifests itself in an increased initial emittance value.

Besides having a larger emittance, the new distribution also differs
from a Gaussian distribution.  This is illustrated in
Fig.~\ref{fig:patternhist}, which shows, for different excitation
patterns, two snapshots of the normalized transverse beam
distributions, the `initial' one at $10^2$~turns and the `final' one
at $10^4$~turns, and how they differ from each other and from a
Gaussian distribution. For instance, for \seventhtp\ (bottom left plot
in Fig.~\ref{fig:patternhist}), simulations indicate a clear shift in
beam population from the core ($< 1\sigma$) towards the regions at
$\pm 1.5\sigma$ and towards the tails at $\pm 4\sigma$.

@@

\begin{figure*}
  \begin{tabular}{cc}
    Reference bunch, no excitation & Excited bunch, maximum amplitude \\
    \includegraphics[width=0.48\textwidth]{profile_v_10thv_slot_50.png} &
    \includegraphics[width=0.48\textwidth]{profile_v_10thv_slot_1300.png}\\
    \end{tabular}
    \caption{Vertical beam profiles measured with the Beam Synchrotron
      Radiation Telescope (BSRT) during the 2016 experiments on
      \tenthtp\ in the vertical plane. The
      transverse damping system was not active on these bunches. The
      beam distributions at the end of the excitation period are shown
      in black in the top plots, with a Gaussian fit in red. The
      bottom plots show the residuals: final profile minus initial
      profile (black); final profile minus its Gaussian fit
      (red). Residuals are expressed as a fraction of the peak profile
      amplitude. The black lines in the plots of the residuals are a
      measure of the overall change of the distribution shape. The red
      lines indicate how different the final distributions are from a
      Gaussian shape. Details of the analysis are given in
      Ref.~\cite{bsrtprofinj}. The distribution of the reference bunch
      (left) is almost unchanged, whereas the bunch experiencing the
      maximum excitation (right) shows a clear shift of particles from
      the axis towards approximately $\pm 2\sigma$.}
    \label{fig:10thexpprof}
\end{figure*}

\begin{figure*}
  \begin{tabular}{cc}
    \includegraphics[width=0.49\textwidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_intensity.png} &
    \includegraphics[width=0.49\textwidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2_rel.png} \\
    \includegraphics[width=0.49\textwidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit1.png} &
    \includegraphics[width=0.49\textwidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2.png} \\
  \end{tabular}	
  \caption{Simulations (with the distribution-tracking method) based
    on the 2016 injection optics with standard lattice errors and
    tunes $(Q_x, Q_y) = (64.28, 59.31)$: relative beam intensity (top
    left); normalized horizontal (bottom left) and vertical (bottom
    right) emittances; vertical emittance relative to its initial
    value (top right). The solid lines indicate the effect of
    \tenthtp\ in the vertical plane. The dotted
    lines include both the resonant excitation and a random dipole
    noise component of 6~nrad (in both horizontal and vertical
    planes).}
  \label{fig:10thsim}
\end{figure*}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{2016_scale_amp_10v_ran_lblshort_sim_prstab.png}
  \caption{Comparison between experiments and simulations of loss
    rates vs.\ resonant excitation amplitude for
    \tenthtp\ in the vertical plane. The
    relative average loss rate~$R$ is defined in
    Eq.~\ref{eqn:lossrate}. The experimental results for a maximum
    excitation amplitude of 48~nrad are shown in black and those for
    96~nrad are plotted in green (see also Fig.~\ref{fig:10thexp},
    left, for instance). The results with damper off are represented
    by solid dots, those with damper on are shown with open
    circles. The simulation results, including random dipole noise,
    are shown in red (see also Fig.~\ref{fig:10thsim}). Statistical
    and systematic uncertainties are discussed in the text. The curves
    represent empirical quadratic fits to the data with damper off
    (solid) and with damper on (dashed).}
  \label{fig:10thexploss}
\end{figure}

The \tenthtp\ pattern was tested in 2016 with an
excitation in the vertical plane only. As this was the first time a
resonant excitation was tested in the LHC, only 48~bunches were
injected, in order to guarantee safe operation of the machine. To test
different excitation amplitudes during one fill and have enough
statistics, the excitation was only applied in one plane. The
experiment started with a period of 12~minutes without excitation, to
let the beam distribution fully adjust to its equilibrium state after
injection. The excitation was then applied for 11~minutes, with the
excitation scheme shown in Fig.~\ref{fig:fill} and a maximum
excitation amplitude of
$A_{\mathrm{max}} = 5 \, \Delta A = \q{48}{nrad}$. The excitation was
then further increased to the maximum value of
$A_{\mathrm{max}} = 5 \, \Delta A = \q{96}{nrad}$ and kept for another
period of 11~minutes.

@@

Numerical simulations of intensities and emittances for this
experiment are shown in Fig.~\ref{fig:10thsim}. They are the results
of distribution tracking for \tenthtp\ without
additional noise (solid lines) and with an additional noise component
(dotted lines), to emulate the natural noise present in the LHC
(Eq.~\ref{eq:noise-kick}). The presence of noise significantly changes
the calculated effect of the resonant excitation on both losses and
emittances.

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{2016+2017_scale_amp_7h_lblshort.png}
  \caption{Measured loss rates vs.\ excitation amplitude for
    \seventhtp\ in the horizontal plane during the 2016 experiment
    (red and yellow) and the 2017 experiment (black and green). The
    uncertainties on the excitation amplitude are dominated by the
    calibration of the transverse feedback and damping system. The
    uncertainties on the loss rates are statistical. An estimate of
    the systematic uncertainties (due to changes in beam distribution,
    for instance) is given by the difference between the two data sets
    within each experiment. The lines indicate empirical second-order
    polynomial fits, with damper off (solid) and with damper on
    (dashed).}
  \label{fig:7thexploss} 
\end{figure}

\begin{figure*}
  \begin{tabular}{ccc}
    \includegraphics[width=0.32\textwidth]{2017_scale_amp_7h_lblshort.png} &
    \includegraphics[width=0.32\textwidth]{2017_scale_amp_7v_lblshort.png} &
    \includegraphics[width=0.32\textwidth]{2017_scale_amp_7hv_lblshort.png} \\
  \end{tabular}
  \caption{Comparison of the measured loss rates as a function of
    excitation amplitude during the 2017 experiment for \seventhtp\ in
    H only (left), V only (center), and in H+V (right). The three
    excitations generate similar loss rates.}
  \label{fig:7thexploss2017}
\end{figure*}

\begin{figure*}
  \begin{tabular}{cc}
    Reference bunch, no excitation & Excited bunch, 7th, H+V \\
    \includegraphics[width=0.48\textwidth]{profile_v_7thhv_slot_2862.png} &
    \includegraphics[width=0.48\textwidth]{profile_v_7thhv_slot_2696.png} \\
  \end{tabular}
  \caption{Vertical beam profiles measured with the Beam Synchrotron
    Radiation Telescope (BSRT) during the 2017 experiments. The
    profiles are taken at the end of \seventhtp\ in H+V. For these
    bunches, the transverse damper was not active. The data is
    presented in the same way as in Fig.~\ref{fig:10thexpprof}. The
    profile changes in the bunch affected by the maximum excitation
    (right) are larger than those displayed by the reference bunch
    (left).}
  \label{fig:7thexpprof}
\end{figure*}

The \seventhtp\ pattern was tested during both the
2016 and 2017 experiments. In 2016, the resonant excitation was
employed for the first time in the LHC and the experiments were
therefore still in an exploratory stage. Based on the experience
gained in 2016, the experiments were then repeated more
systematically, with more bunches and on both excitation planes.
